The CrypTool Script: \\ Cryptography, Mathematics and More \\ \vspace{30pt} {\large Background reading\\%*[-8pt] for the free e-learning program CrypTool\\ (10th edition -- distributed with CrypTool version 1.4.30)} \vspace{60pt}

### 2009-08-03 script/en/chapters/numbertheory.tex
CT script: added new cover page (de+en); chap 4: appendix with sage samples: changed the name of some self-defined functions... - not finished yet (en only)

\begin{sagecode}
\begin{Verbatim}%
[fontsize=\footnotesize,fontshape=tt]
sage: def enum_primroots(M):
....:     r"""
....:     Return all the primitive roots of M.
....:     """
....:     g = primitive_root(M)
....:     targetOrder = euler_phi(M)
....:     L = []
....:     # Stepping through all odd integers from 1 up to M, not including
....:     # M. So this loop only considers values of i where 1 <= i < M.
....:     for i in xrange(1, M, 2):
....:         testGen = Mod(g^i, M)
....:         if testGen.multiplicative_order() == targetOrder:
....:             L.append(testGen.lift())
....:     # removing duplicates
....:     return Set(L)
\end{Verbatim}
\caption{Function to calculate all primitive roots for a given number}
\end{sagecode}

@@

\begin{sagecode}
\begin{Verbatim}%
[fontsize=\footnotesize,fontshape=tt]
sage: enum_primroots(541)
{2, 517, 10, 523, 13, 14, 527, 528, 18, 531, 24, 539, 30, 37, 40, 51,
54, 55, 59, 62, 65, 67, 68, 72, 73, 77, 83, 86, 87, 91, 94, 96, 98,
99, 107, 113, 114, 116, 117, 126, 127, 128, 131, 132, 138, 150, 152,
153, 156, 158, 163, 176, 181, 183, 184, 195, 197, 199, 206, 208,
210, 213, 218, 220, 223, 224, 244, 248, 250, 257, 258, 259, 260,
261, 263, 267, 269, 270, 271, 272, 274, 278, 280, 281, 282, 283,
284, 291, 293, 297, 317, 318, 321, 323, 328, 331, 333, 335, 342,
344, 346, 357, 358, 360, 365, 378, 383, 385, 388, 389, 391, 403,
409, 410, 413, 414, 415, 424, 425, 427, 428, 434, 442, 443, 445,
447, 450, 454, 455, 458, 464, 468, 469, 473, 474, 476, 479, 482,
486, 487, 490, 501, 504, 511}
\end{Verbatim}
\caption{Table with all primitive roots for a given prime}
\end{sagecode}

%########## Find number of primes and primitive roots in given range and time
\noindent With a little bit of programming, we can find out how long
it takes to compute all the primitive roots of some number of primes.

\begin{sagecode}
\begin{Verbatim}%
[fontsize=\footnotesize,fontshape=tt]
%a##########################
sage: def primroots_iter_tell_primenumber(start, end):
....:      r"""
....:      Compute all primitive roots of all primes between start and end,
....:      inclusive. This uses a primes iterator.
....:      """
....:      nPrimes = 0
....:      nPrimitiveRoots = 0
....:      for p in primes(start, end+1):
....:              nPrimes += 1
....:          L = enum_primroots(p)
....:          nPrimitiveRoots += len(L)
....:      print "Enumerate all %s" % nPrimitiveRoots + " primitive roots of %s primes." % nPrimes
....:
sage: time primroots_iter_tell_primenumber(1, 1500)
Enumerate all 62044 primitive roots of 239 primes.
CPU times: user 6.46 s, sys: 0.18 s, total: 6.64 s
Wall time: 7.48 s

sage: def primroots_iter(start, end):
....:     r"""
....:     Compute all primitive roots of all primes between start and end,
....:     inclusive. This uses a primes iterator.
....:     """
....:     for p in primes(start, end+1):
....:         enum_primroots(p)
....:         
sage: %time primroots_iter(1, 1500)
CPU times: user 2.04 s, sys: 0.17 s, total: 2.21 s
Wall time: 2.21 s
\end{Verbatim}
\caption{Function to calculate all primitive roots for some given primes}
\end{sagecode}

@@

%a##########################
\begin{sagecode}
\begin{Verbatim}%
[fontsize=\footnotesize,fontshape=tt]
sage: def primroots_counter(start, end, bPrimesOnly=True):
	r"""
	Compute all primitive roots of all primes between start and end,
	inclusive. This actually performs primality tests, so it allows us
	to count the number of primes from start to end, inclusive.
	"""
	nCheckedNumb = 0
	nPrimitiveRoots = 0
	for n in xrange(start, end+1):
		if bPrimesOnly:
			if is_prime(n):
				nCheckedNumb += 1
				L = enum_primroots(n)
				nPrimitiveRoots += len(L)
		else:
			nCheckedNumb += 1
			L = enum_primroots(n)
			nPrimitiveRoots += len(L)
	if bPrimesOnly:
		print "Enumerate all %s" % nPrimitiveRoots + " primitive roots of %s primes." % nCheckedNumb
	else:
		print "Enumerate all %s" % nPrimitiveRoots + " primitive roots of %s numbers." % nCheckedNumb

>>>>

\begin{sagecode}
\begin{Verbatim}%
[fontsize=\footnotesize,fontshape=tt]
sage: def enum_PrimitiveRoots_of_an_Integer(M):
....:     r"""
....:     Return all the primitive roots of M.
....:     """
....:     g = primitive_root(M)
....:     targetOrder = euler_phi(M)
....:     L = []
....:     # Stepping through all odd integers from 1 up to M, not including
....:     # M. So this loop only considers values of i where 1 <= i < M.
....:     for i in xrange(1, M, 2):
....:         testGen = Mod(g^i, M)
....:         if testGen.multiplicative_order() == targetOrder:
....:             L.append(testGen.lift())
....:     # removing duplicates
....:     return Set(L)

###aa
Alternativ-Code: In Def. abfangen, wenn die Fkt. primitive_root nicht berechenbar. Tut noch nicht.
sage: def enum_PrimitiveRoots_of_an_Integer(M):
	r"""
	Return all the primitive roots of M.
	"""
	try:
		g = primitive_root(M)
		print "Das ist %s" % M
	except:
		# raise ArithmeticError, "primitive_root() failed for %s." % M
		raise ArithmeticError, "primitive_root() failed."
	targetOrder = euler_phi(M)
	L = []
	# Stepping through all odd integers from 1 up to M, not including
	# M. So this loop only considers values of i where 1 <= i < M.
	for i in xrange(1, M, 2):
		testGen = Mod(g^i, M)
		if testGen.multiplicative_order() == targetOrder:
			L.append(testGen.lift())
	# removing duplicates
	return Set(L)

###zz

\end{Verbatim}
\caption{Function to calculate all primitive roots for a given number}
\end{sagecode}

@@

\begin{sagecode}
\begin{Verbatim}%
[fontsize=\footnotesize,fontshape=tt]
sage: L=enum_PrimitiveRoots_of_an_Integer(541)
{2, 517, 10, 523, 13, 14, 527, 528, 18, 531, 24, 539, 30, 37, 40, 51,
54, 55, 59, 62, 65, 67, 68, 72, 73, 77, 83, 86, 87, 91, 94, 96, 98,
99, 107, 113, 114, 116, 117, 126, 127, 128, 131, 132, 138, 150, 152,
153, 156, 158, 163, 176, 181, 183, 184, 195, 197, 199, 206, 208,
210, 213, 218, 220, 223, 224, 244, 248, 250, 257, 258, 259, 260,
261, 263, 267, 269, 270, 271, 272, 274, 278, 280, 281, 282, 283,
284, 291, 293, 297, 317, 318, 321, 323, 328, 331, 333, 335, 342,
344, 346, 357, 358, 360, 365, 378, 383, 385, 388, 389, 391, 403,
409, 410, 413, 414, 415, 424, 425, 427, 428, 434, 442, 443, 445,
447, 450, 454, 455, 458, 464, 468, 469, 473, 474, 476, 479, 482,
486, 487, 490, 501, 504, 511}
sage: L
sage: len(L)
144
\end{Verbatim}
\caption{Table with all primitive roots for a given prime}
\end{sagecode}

%########## Find number of primes and primitive roots in given range and time
\noindent With a little bit of programming, we can count how many primes are in a given range
and enumerate all their primitive roots.

With time command we can find out how long this takes.

\begin{sagecode}
\begin{Verbatim}%
[fontsize=\footnotesize,fontshape=tt]
sage: def enum_PrimitiveRoots_of_a_Range_of_Primes(start, end):
....:      r"""
....:      Compute all primitive roots of all primes between start and end,
....:      inclusive. This uses a primes iterator.
....:      """
....:      nPrimes = 0
....:      nPrimitiveRoots = 0
....:      for p in primes(start, end+1):
....:              nPrimes += 1
....:          L = enum_PrimitiveRoots_of_an_Integer(p)
....:          nPrimitiveRoots += len(L)
....:      print "Enumerate all %s" % nPrimitiveRoots + " primitive roots of %s primes." % nPrimes
....:
sage: time primroots_iter_tell_primenumber(1, 1500)
Enumerate all 62044 primitive roots of 239 primes.
CPU times: user 6.46 s, sys: 0.18 s, total: 6.64 s
Wall time: 7.48 s
\end{Verbatim}
\caption{Function to calculate all primitive roots for some given primes}
\end{sagecode}

@@

%a##########################
\begin{sagecode}
\begin{Verbatim}%
[fontsize=\footnotesize,fontshape=tt]
sage: def enum_PrimitiveRoots_of_a_Range_of_Integers(start, end, bPrimesOnly=True):
	r"""
	Compute all primitive roots of all primes between start and end,
	inclusive. This actually performs primality tests, so it allows us
	to count the number of primes from start to end, inclusive.
	"""
	nCheckedNumb = 0
	nPrimitiveRoots = 0
	for n in xrange(start, end+1):
		if bPrimesOnly:
			if is_prime(n):
				nCheckedNumb += 1
				L = enum_PrimitiveRoots_of_an_Integer(n)
				nPrimitiveRoots += len(L)
		else:
			nCheckedNumb += 1
			L = enum_PrimitiveRoots_of_an_Integer(n)
			nPrimitiveRoots += len(L)
	if bPrimesOnly:
		print "Enumerate all %s" % nPrimitiveRoots + " primitive roots of %s primes." % nCheckedNumb
	else:
		print "Enumerate all %s" % nPrimitiveRoots + " primitive roots of %s numbers." % nCheckedNumb
